Combined annotation exploration for SCPCP000006

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Some text cleanup and OSPL

In a second time, we will explore and validate the chosen annotation.

@@

-   `pred.thr` = `r params$predicted.celltype.threshold`

-   `cnv.thr.lower` = `r params$cnv_threshold_low`

-   `cnv.thr.upper` = `r params$cnv_threshold_high`

@@

In this notebook, we are working with all of the samples in SCPCP000006.

The sample metadata can be found in `sample_metadata_file` in the `data` folder.

@@

We extracted from the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`. the following information per cell:

-   the predicted compartment in `fetal_kidney_predicted.compartment` and `fetal_kidney_predicted.compartment.score`

-   the predicted cell type in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`

-   the sample identifier in `sample_id`

-   the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22

-   the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)

@@

-   the samples metadata or clinical data reported in `r file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")`

```{r }
# Add clinical metadata
cell_type_df <- cell_type_df |>
  left_join(metadata, by = c("sample_id" = "scpca_sample_id"))
```

-   the cytoband file to get the length of each chromosome arm (for plotting only)

@@

We simply checked for each chromosome the cell `cnv_chr`. Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `cnv_score` will be FALSE. Would the cell have more than `r params$cnv_threshold_high` chromosome with CNV, the global `cnv_score` will be TRUE. The default is set to `unknown`.

@@

-   *normal* cells can be observe in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV. We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score. Indeed, we know that false positive CNV can be observed in a cell type specific manner. Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` \> `r params$predicted.celltype.threshold` as `normal`.

@@

-   *cancer* cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV

@@

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` *cancer* cells, `r table(cell_type_df$first.level_annotation)["normal"]` *normal* cells. `r table(cell_type_df$first.level_annotation)["unknown"]` cells remain *unknown*

@@

These five samples do not have enough immune and endothelium cell to be used as a normal reference in `scripts/06_infercnv.R` and we previously decided to run `scripts/06_infercnv.R` without any reference for these samples. In that case, the mean expression profile over the sample is taken as the normal reference, biasing the inference of CNV.

For those samples, the annotations based on `infercnv` cannot be trust. To avoid any confusion in the following steps of the analysis, we decided to force the annotation of the entire samples to `unknown`.

@@

-   Normal cells from the `fetal nephron` compartment must be normal kidney cells.

-   Normal cells from the `stroma` compartment must be normal stroma cells.

-   Immune and endothelial cells have been already identified by label transfer.

@@

-   *cancer stroma*: We define as *cancer stroma* all cancer cells from the stroma compartment.

-   *blastema*,: we defined as *blastema* every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`. We know that these *mesenchymal* cells are cells from the cap mesenchyme that are not expected to be in a mature kidney. These blastema cells should express higher `CITED1` and/or `NCAM1`.

-   *cancer epithelium*: we defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells. We expect these cells to express epithelial markers. Their predicted cell type should correspond to more mature kidney epithelial subunits.

@@

-   validate that mostly tumor cells do present high percentages of cells with CNV (by definition)

-   identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as the *chr6p* which is a hub for immune genes

-   get the repartition per sample of the CNV

@@

While the above heatmaps can be informative, we have to acknowledge that there are numerous and it is difficult to get a global picture of the CNV profile of the entire dataset.

Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). This represent the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).

@@

-   *chr8q*, *chr12*, *chr13*, *chr18q* gains

-   *chr11q*, *chr16*, *chr17q* loss

@@

-   we couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

-   we identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). These can be `infercnv` induced \_FALSE_positive CNV.

@@

Blastema cells are enriched in *CITED1*+ cells, but some samples do not express it at all.

>>>>

We will then explore and validate the chosen annotation.

@@

- `pred.thr` = `r params$predicted.celltype.threshold`
- `cnv.thr.lower` = `r params$cnv_threshold_low`
- `cnv.thr.upper` = `r params$cnv_threshold_high`

@@

In this notebook, we are working with all of the samples in SCPCP000006.
The sample metadata can be found in `sample_metadata_file` in the `data` folder.

@@

From the pre-processed and labeled `Seurat` object from `results/{sample_id}/06_infercnv_HMM-i3_{sample_id}_reference-{reference}.rds`, we will extract the following information per cell:

- the predicted compartment in `fetal_kidney_predicted.compartment` and `fetal_kidney_predicted.compartment.score`
- the predicted cell type in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`
- the sample identifier in `sample_id`
- the CNV `{dupli|loss}` estimation per chromosome {i} and arm `{p|q}`: `has_{dupli|loss}_chr{i}{p|q}`, in 1 to 22
- the `counts` of [markers genes](https://github.com/AlexsLemonade/OpenScPCA-analysis/tree/main/analyses/cell-type-wilms-tumor-06#the-table-celltype_metadatacsv-contains-the-following-column-and-information)

@@

- the samples metadata or clinical data reported in `r file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")`
- the cytoband file to get the length of each chromosome arm (for plotting only)

```{r metadata}
# Add clinical metadata
cell_type_df <- cell_type_df |>
  left_join(metadata, by = c("sample_id" = "scpca_sample_id"))
```

@@

We simply checked the `cnv_chr` for each chromosome across cells.

- If the cell has less than `r params$cnv_threshold_low` chromosome with CNV, the global `cnv_score` will be `FALSE`. 
- If the cell has more than `r params$cnv_threshold_high` chromosome with CNV, the global `cnv_score` will be `TRUE`. 
- The default is set to `unknown`.

@@

- *normal* cells can be observed in all five compartments (`normal`,`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV. 
- *cancer* cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV

We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score. 
Indeed, we know that false positive CNV can be observed in a cell type specific manner. 
Please note that in `06_infercnv.R`, we renamed all `immune` and `endothelium` cells with a `fetal_kidney_predicted.compartment.score` \> `r params$predicted.celltype.threshold` as `normal`.

@@

@@

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` *cancer* cells, `r table(cell_type_df$first.level_annotation)["normal"]` *normal* cells.

`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain *unknown*.

@@

These five samples do not have enough immune and endothelium cell to be used as a normal reference in `scripts/06_infercnv.R` and we previously decided to run `scripts/06_infercnv.R` without any reference for these samples. 
Without a specified reference, the mean expression profile across the sample is taken as the normal reference, biasing the inference of CNV.
For those samples, the annotations based on `infercnv` cannot be trusted. 
To avoid any confusion in the following steps of the analysis, we decided to force the annotation of the entire samples to `unknown`.

@@

- Normal cells from the `fetal nephron` compartment must be normal kidney cells.
- Normal cells from the `stroma` compartment must be normal stroma cells.
- Immune and endothelial cells have been already identified by label transfer.

@@

- *cancer stroma*: We define as *cancer stroma* all cancer cells from the stroma compartment.
- *blastema*: We defined as *blastema* every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`. 
We know that these *mesenchymal* cells are cells from the cap mesenchyme that are not expected to be in a mature kidney. 
These blastema cells should express higher `CITED1` and/or `NCAM1`.
- *cancer epithelium*: We defined as *cancer epithelium* all cancer cells that are neither stroma nor blastemal cells. 
We expect these cells to express epithelial markers. 
Their predicted cell type should correspond to more mature kidney epithelial subunits.

@@

- validate that mostly tumor cells do present high percentages of cells with CNV (by definition)
- identify chromosome and cell types that might be more sensitive to *FALSE* positive CNV, such as `chr6p` which is a hub for immune genes
- get the repartition per sample of the CNV

@@

While the above heatmaps can be informative, because there are so many of them it is difficult to get a global picture of the CNV profile of the entire dataset.

Here we aim to plot a mean CNV profile across the 35 annotated samples as presented in Figure 1A of [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
This represents the mean SNP Array profile of 64 pediatric kidney cancer (including 60 Wilms tumors).

@@

- *chr8q*, *chr12*, *chr13*, *chr18q* gains
- *chr11q*, *chr16*, *chr17q* loss

@@

- We couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

- We identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
These may be `infercnv` induced false-positive CNV.

@@

Blastema cells are enriched in `CITED1+` cells, but some samples do not express it at all.
